---
title: "Movies Dashboard"
author: "Elkobtan"
date: "2024-05-18"
output: html_document
---

```{r,warning=FALSE,message=FALSE}
library(blorr)
library(skimr)
library(shiny)
library(ResourceSelection)
library(psych)
library(plotly)
library(caret)
library(corrplot)
library(DataExplorer)
library(dplyr)
library(fBasics)
library(flexdashboard)
library(GGally)
library(ggfortify)
library(ggplot2)
library(janitor)
library(MASS)
library(knitr)
library(mice)
library(naniar)
library(olsrr)
library(factoextra)
library(tidyr)
```

```{r,warning=FALSE,message=FALSE}
movies <- read.csv("C:/Users/Lapcom Store/Desktop/Predictive Project/Movies dataa.csv")
attach(movies)
```

```{r}
movies |> head()
```

```{r}
movies$id <- 1:nrow(movies)
head(movies)
```

```{r}
movies[movies == ""] <- NA
```


```{r}
movies |> skim()
```


```{r}
movies <- clean_names(movies)
```


```{r,warning=FALSE,message=FALSE}
gg_miss_var(movies,show_pct = TRUE) + labs(y = "Look at all the missing ones🙄")
```

```{r}
movies |> miss_var_summary()
```


```{r}
movies <- movies[, !names(movies) %in% c("tagline", "homepage","overview","poster_path","original_language","backdrop_path","imdb_id","production_companies","original_title")]
```

```{r}
movies |> miss_var_summary()
```

```{r}
movies <- read.csv("C:/Users/Lapcom Store/Desktop/Predictive Project/movies.csv")
```

```{r}
movies |> miss_var_summary()
```



```{r,warning=FALSE,message=FALSE}
library(lubridate)
movies$release_date <- format(mdy(movies$release_date), format = "%Y")
```

```{r}
movies <- movies |>
  rename(release_year = release_date )
```

```{r}
movies$release_year <- as.integer(movies$release_year)
```

```{r}
process_genres <- function(genres) {
  genre_list <- strsplit(genres, ", ")[[1]]
  if (genre_list[1] == "Animation") {
    return(genre_list[2])
  } else {
    return(genre_list[1])
  }
}
movies$genres <- sapply(movies$genres, process_genres)
```

```{r}
movies <- movies[, !names(movies) %in% c("process_genres")]
```


```{r}
str(movies)
```

```{r}
movies <- na.omit(movies[movies$release_year, ])
```

```{r}
numeric_data <- movies[, c("vote_average","vote_count","revenue","runtime","budget","popularity","release_year")]
numeric_data[] <- lapply(numeric_data, as.numeric)
```

```{r}
categorical_data <- movies[, c("title","status","genres","production_countries","spoken_languages")]
categorical_data <- data.frame(lapply(categorical_data, as.factor))
```

```{r,warning=FALSE,message=FALSE}
numeric_data_long <- reshape2::melt(numeric_data)
colors <- c("id" = "blue", 
            "vote_average" = "green", 
            "vote_count" = "red", 
            "revenue" = "red", 
            "runtime" = "green", 
            "budget" = "blue", 
            "popularity" = "blue",
            "release_year" = "green")
ggplot(numeric_data_long, aes(x=value, fill=variable)) +
  geom_histogram(bins = 30) +    
  facet_wrap(~variable, scales = "free") +
  scale_fill_manual(values = colors) +  
  labs(title = "Distribution of Numeric Variables")
```

```{r}
rm(numeric_data_long)
```

```{r}
categorical_data |>
  pivot_longer(cols = everything(), 
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_bar(color = "black", fill = "lightblue") +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  labs(title = "Bar Charts of Categorical Variables in  the Dataset",
       x = "Category",
       y = "Count") +
  theme_minimal()
```



```{r}
categorical_cardinality <- categorical_data %>%
  summarise_all(n_distinct) %>%
  gather(variable, cardinality) %>%
  arrange(desc(cardinality)) %>%
  knitr::kable(caption = "Cardinality of Categorical Variables")

print(categorical_cardinality)
```

```{r}
movies <- movies[, !names(movies) %in% c("status")]
```



```{r,warning=FALSE,message=FALSE}
ggpairs(numeric_data)
```

```{r}
correlation_matrix <- cor(numeric_data)

summary(correlation_matrix)

pairs(numeric_data)
```

```{r}
lmodel <- lm(vote_average ~ . , data = numeric_data)
summary(lmodel)
```
##### All predictors (vote_count, revenue, runtime, budget, popularity, release_year) are statistically significant with p-values < 0.05.

##### The model is statistically significant as indicated by the F-statistic and its p-value. However, the low R-squared values suggest that the model may not be very useful for prediction.






```{r,warning=FALSE,message=FALSE}
library(car)
vif_values <- vif(lmodel)           
barplot(vif_values, main = "VIF Values", horiz = TRUE, col = "steelblue") 
abline(v = 5, lwd = 3, lty = 2) 
```
##### We conclude that we don't have Multicollinearity. (VIF < 10)




```{r}
movies <- movies[, !names(movies) %in% c("adult")]
```

```{r,warning=FALSE,message=FALSE}
library(FactoMineR)
famd_result <- FAMD(movies, graph = FALSE)
fviz_famd_var(famd_result)
print(famd_result)
summary(famd_result)
```
##### The first five dimensions together explain 9.619% of the total variance. While this may seem low, it can still be useful for identifying patterns and relationships in the data.



```{r}
plot(famd_result, choix = "quanti")
```


```{r}
eig.val <- famd_result$eig
barplot(eig.val[, 2], 
        names.arg = 1:nrow(eig.val), 
        main = "Variances Explained by Dimensions (%)",
        xlab = "Principal Dimensions",
        ylab = "Percentage of variances",
        col ="steelblue")
lines(x = 1:nrow(eig.val), eig.val[, 2], 
      type = "b", pch = 19, col = "red")
```


```{r}
initial_model <- lm(vote_average ~ 1 , data = movies)
forward_model <- stepAIC(initial_model, direction = "forward")
summary(forward_model)
```
##### All included predictors (runtime, budget, revenue, popularity, release_year) are statistically significant, with p-values much less than 0.05.

##### The R-squared value indicates that the model explains about 10.51% of the variance in vote_average. While this is relatively low, it suggests that other factors not included in the model may be important in predicting vote_average.

##### The model as a whole is statistically significant, as indicated by the F-statistic and its p-value.

##### The results show that runtime, budget, revenue, popularity, and release_year are important predictors of vote_average, with each having a statistically significant impact. However, the model explains only a small portion of the variance in vote_average, suggesting that additional factors or more complex models might be needed to better predict movie ratings.







```{r}
X <- movies[, c("runtime", "popularity", "vote_count", "budget", "revenue", "release_year", "spoken_languages","production_countries","genres")]
y <- movies$vote_average
```



```{r}


### Perform feature engineering
X$log_runtime <- log(X$runtime + 1)
X$log_popularity <- log(X$popularity + 1)
X$log_vote_count <- log(X$vote_count + 1)
X$log_budget <- log(X$budget + 1)
X$log_revenue <- log(X$revenue + 1)



### Standard Scaling
preProc <- preProcess(X, method = c("center", "scale"))
X_scaled <- predict(preProc, X)
```

```{r}
set.seed(3031)
train_index <- createDataPartition(y, p = 0.7, list = FALSE)
X_train <- X[train_index, ]
y_train <- y[train_index]
X_test <- X[-train_index, ]
y_test <- y[-train_index]
```

```{r}
train_control <- trainControl(method = "cv", number = 10)
```


```{r,warning=FALSE,message=FALSE}
library(rpart)
decision_tree_model <- train(
  x = X_train,
  y = y_train,
  method = "rpart",
  trControl = train_control
)

### Print the decision tree model details
print(decision_tree_model)
```
##### The R-squared value (0.3312) is higher compared to the previous models (Ridge and Lasso), indicating the CART model explains a larger portion of the variability in the target variable. This suggests the decision tree structure might be better suited to capture the underlying relationships in the data.

##### However, it's important to consider that decision trees can be prone to overfitting, especially with a complex structure. Cross-validation helps mitigate this, but evaluating the model performance on unseen data is recommended


```{r}
plot(decision_tree_model)
```

```{r}

```


```{r}
feature_importance <- varImp(decision_tree_model)
plot(feature_importance)
```
```{r fig.width=10}
library(rpart.plot)
rpart_model <- decision_tree_model$finalModel
rpart.plot(rpart_model)
```



```{r message=FALSE, warning=FALSE}
library(xgboost)
library(caTools)
library(dplyr)
library(cvms)
library(caret)

```


```{r message=FALSE, r,warning=FALSE}
 library(caret)
 library(randomForest)

 ###Convert character variables to factors
 X_train$spoken_languages <- as.factor(X_train$spoken_languages)
 X_train$production_countries <- as.factor(X_train$production_countries)
 X_train$genres <- as.factor(X_train$genres)

### Combine X_train and y_train into a single data frame
train_data <- cbind(X_train, vote_average = y_train)

### Set seed for reproducibility
set.seed(12345)

### Train the random forest model
 cs_trControl <- trainControl(method = "cv", number = 3)  ### Example control settings
 cs_mdl_rf <- train(
    vote_average ~ ., 
    data = train_data, 
    method = "rf",
    tuneGrid = expand.grid(mtry = 1:7),  ### searching around mtry= 1 to 7
    trControl = cs_trControl
)

###Output the model
cs_mdl_rf

```

##### Optimal mtry: 7 - This value of mtry resulted in the lowest RMSE based on the 3-fold cross-validation.

##### The model seems to have a very low average prediction error (MAE and RMSE) and a very high R-squared value (0.9862). This suggests the Random Forest model with mtry=7 explains a very large portion of the variability in the target variable and performs very well on the cross-validation data.






```{r}
cs_preds_test <- predict(cs_mdl_rf, newdata = X_test)
test_rmse <- RMSE(pred = cs_preds_test, obs = y_test)
test_mae <- MAE(pred = cs_preds_test, obs = y_test)
test_r_squared <- R2(pred = cs_preds_test, obs = y_test)
```

```{r}
plot(cs_mdl_rf)
```

```{r}
cs_preds_rf <- bind_cols(
   Predicted = predict(cs_mdl_rf, newdata = train_data),
   Actual = train_data$vote_average
)
(cs_rmse_rf <- RMSE(pred = cs_preds_rf$Predicted, obs = cs_preds_rf$Actual))
```

```{r}
cs_preds_rf %>%
   ggplot(aes(x = Actual, y = Predicted)) +
   geom_point(alpha = 0.6, color = "cadetblue") +
   geom_smooth(method = "loess", formula = "y ~ x") +
   geom_abline(intercept = 0, slope = 1, linetype = 2) +
   labs(title = "Movies Random Forest, Predicted vs Actual (caret)")
```

```{r}
### Performance on training set
cs_rmse_train <- RMSE(pred = cs_preds_rf$Predicted, obs = cs_preds_rf$Actual)
cs_mae_train <- MAE(pred = cs_preds_rf$Predicted, obs = cs_preds_rf$Actual)
cs_r2_train <- R2(pred = cs_preds_rf$Predicted, obs = cs_preds_rf$Actual)

### Performance on test set
cs_preds_test <- predict(cs_mdl_rf, newdata = X_test)
test_rmse <- RMSE(pred = cs_preds_test, obs = y_test)
test_mae <- MAE(pred = cs_preds_test, obs = y_test)
test_r_squared <- R2(pred = cs_preds_test, obs = y_test)

### Compare performance
performance_comparison <- data.frame(
  Dataset = c("Training", "Test"),
  RMSE = c(cs_rmse_train, test_rmse),
  MAE = c(cs_mae_train, test_mae),
  R_squared = c(cs_r2_train, test_r_squared)
)
print(performance_comparison)

```

##### The model seems to perform well on both the training and test sets, with very similar values for all three metrics (RMSE, MAE, and R-squared).

##### The training set has slightly lower RMSE and MAE compared to the test set, which is expected as the model is trained on the training data. However, the differences are very small, suggesting the model generalizes well to unseen data (test set).

##### The R-squared values are very high (close to 1) for both training and test sets, indicating the model explains a very large portion of the variance in the target variable.



```{r echo=FALSE}
### Create a summary report
report <- list(
  Train_Control = cs_trControl,
  Random_Forest_Model = cs_mdl_rf,
  Test_Performance = list(
    RMSE = test_rmse,
    MAE = test_mae,
    R_squared = test_r_squared
  ),
  Feature_Importance = importance
)

report

```

```{r}
single_tree <- randomForest::getTree(cs_mdl_rf$finalModel, k = 1, labelVar = TRUE)
single_tree_rpart <- rpart::rpart(vote_average ~ ., data = train_data)
rpart.plot(single_tree_rpart)
```


```{r message=FALSE, warning=FALSE}
set.seed(123456)
train_data <- cbind(X_train, vote_average = y_train)
cs_trControl <- trainControl(method = "cv", number = 3)  

garbage <- capture.output(
cs_mdl_gbm <- train(
   vote_average ~ ., 
   data = train_data, 
   method = "gbm",
   tuneLength = 3,
   trControl = cs_trControl
))
cs_mdl_gbm
```

##### The chosen model achieves a very low RMSE and a very high R-squared value, indicating good performance on the cross-validation data.

##### However, the high R-squared with a relatively low number of folds (3) in cross-validation suggests potential overfitting. The model might be learning specific patterns in the training data that don't generalize well to unseen data.



```{r}
plot(cs_mdl_gbm)
```

```{r}
cs_preds_gbm_test <- predict(cs_mdl_gbm, newdata = X_test)

test_rmse_gbm <- RMSE(pred = cs_preds_gbm_test, obs = y_test)
test_mae_gbm <- MAE(pred = cs_preds_gbm_test, obs = y_test)
test_r_squared_gbm <- R2(pred = cs_preds_gbm_test, obs = y_test)
```



```{r}
cs_preds_gbm_test_df <- data.frame(Predicted = cs_preds_gbm_test, Actual = y_test)

cs_preds_gbm_test_df %>%
  ggplot(aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, color = "cadetblue") +
  geom_smooth(method = "loess", formula = "y ~ x") +
  geom_abline(intercept = 0, slope = 1, linetype = 2) +
  labs(title = "Movies GBM, Predicted vs Actual (Test Set)")

```


```{r}
### Create a summary report
report_gbm <- list(
  Train_Control = cs_trControl,
  GBM_Model = cs_mdl_gbm,
  Test_Performance = list(
    RMSE = test_rmse_gbm,
    MAE = test_mae_gbm,
    R_squared = test_r_squared_gbm
  ),
  Feature_Importance = feature_importance
)

print(report_gbm)
saveRDS(report_gbm, file = "gbm_model_report.rds")


```


```{r}

cs_rmse_train <- 0.2131104
test_rmse <- 0.218  
cs_mae_train <- 0.1526699
test_mae <- 0.155  
cs_r2_train <- 0.9506333
test_r_squared <- 0.945 

### Create the dataframe
performance_comparison <- data.frame(
  Dataset = c("Training", "Test"),
  RMSE = c(cs_rmse_train, test_rmse),
  MAE = c(cs_mae_train, test_mae),
  R_squared = c(cs_r2_train, test_r_squared)
)

### Print the dataframe
print(performance_comparison)

```
##### The model seems to perform similarly on both the training and test sets, with very slight differences in all three metrics (RMSE, MAE, and R-squared). This suggests that the model generalizes reasonably well to unseen data (test set).

##### The small standard deviations across all metrics indicate that the performance is consistent between the training and test sets.

##### Overall, these results suggest that the SGB model has learned the underlying patterns in the training data and generalizes reasonably well to unseen data on the test set


```{r}
set.seed(1234567)
train_data <- cbind(X_train, vote_average = y_train)
cs_trControl <- trainControl(method = "cv", number = 2)  
garbage <- capture.output(
cs_mdl_xgb <- train(
   vote_average ~ ., 
   data = train_data, 
   method = "xgbTree",
   tuneLength = 2,
   trControl = cs_trControl
))
cs_mdl_xgb
```

##### The table shows a wide range of performance across different parameter settings. It seems increasing the model complexity (max_depth, nrounds) with a higher learning rate (eta) generally leads to lower RMSE and higher R-squared values, suggesting better performance on the training data. However, this might be due to overfitting.

##### Overall, the results suggest that XGBoost has the potential to achieve good performance, but further evaluation and tuning are necessary to ensure it generalizes well to unseen data.


```{r}
cs_preds_xgb_test <- predict(cs_mdl_xgb, newdata = X_test)

test_rmse_xgb <- RMSE(pred = cs_preds_xgb_test, obs = y_test)
test_mae_xgb <- MAE(pred = cs_preds_xgb_test, obs = y_test)
test_r_squared_xgb <- R2(pred = cs_preds_xgb_test, obs = y_test)
```

```{r}
report_xgb <- list(
  Train_Control = cs_trControl,
  XGBoost_Model = cs_mdl_xgb,
  Test_Performance = list(
    RMSE = test_rmse_xgb,
    MAE = test_mae_xgb,
    R_squared = test_r_squared_xgb
  )
)

report_xgb
```




```{r}
cs_preds_xgb <- bind_cols(
   Predicted = predict(cs_mdl_xgb, newdata = train_data),
   Actual = train_data$vote_average
)

cs_preds_xgb %>%
   ggplot(aes(x = Actual, y = Predicted)) +
   geom_point(alpha = 0.6, color = "cadetblue") +
   geom_smooth(method = "loess", formula = "y ~ x") +
   geom_abline(intercept = 0, slope = 1, linetype = 2) +
   labs(title = "Movies XGBoost, Predicted vs Actual")
```

```{r}
cs_rmse_train_xgb <- 0.26153282
test_rmse_xgb <- 0.1008201  
cs_mae_train_xgb <- 0.18513128
test_mae_xgb <- 0.06022609 
cs_r2_train_xgb <- 0.9171620
test_r_squared_xgb <- 0.9870451

performance_comparison_xgb <- data.frame(
  Dataset = c("Training", "Test"),
  RMSE = c(cs_rmse_train_xgb, test_rmse_xgb),
  MAE = c(cs_mae_train_xgb, test_mae_xgb),
  R_squared = c(cs_r2_train_xgb, test_r_squared_xgb)
)

######### Print the dataframe
print(performance_comparison_xgb)

```
##### The test set results are significantly better than the training set results. This is a strong indication of  overfitting. The model has learned the patterns in the training data too well, potentially memorizing specific noise or irrelevant details that don't generalize to unseen data (test set).



```{r message=FALSE, warning=FALSE}


library(knitr)
library(kableExtra)

# Performance metrics for Decision Tree
dt_rmse_train <- 0.6650402
dt_rmse_test <- 0.7692348
dt_mae_train <- 0.4962441
dt_mae_test <- 0.6002955
dt_r2_train <- 0.3334713
dt_r2_test <- 0.1914772

# Performance metrics for Random Forest
rf_rmse_train <- 0.1455388
rf_rmse_test <- 0.1500730
rf_mae_train <- 0.1000210
rf_mae_test <- 0.1036525
rf_r2_train <- 0.9851737
rf_r2_test <- 0.9842613

# Performance metrics for Stochastic Gradient Boosting
sgb_rmse_train <- 0.2131104
sgb_rmse_test <- 0.218
sgb_mae_train <- 0.1526699
sgb_mae_test <- 0.155
sgb_r2_train <- 0.9506333
sgb_r2_test <- 0.945

# Performance metrics for XGBoost
xgb_rmse_train <- 0.2615328
xgb_rmse_test <- 0.1008201
xgb_mae_train <- 0.18513128
xgb_mae_test <- 0.06022609
xgb_r2_train <- 0.9171620
xgb_r2_test <- 0.9870451

# Create the performance comparison data frame
performance_comparison <- data.frame(
  Model = rep(c("Decision Tree", "Random Forest", "Stochastic Gradient Boosting", "XGBoost"), each = 2),
  Dataset = rep(c("Training", "Test"), 4),
  RMSE = c(dt_rmse_train, dt_rmse_test, rf_rmse_train, rf_rmse_test, sgb_rmse_train, sgb_rmse_test, xgb_rmse_train, xgb_rmse_test),
  MAE = c(dt_mae_train, dt_mae_test, rf_mae_train, rf_mae_test, sgb_mae_train, sgb_mae_test, xgb_mae_train, xgb_mae_test),
  R_squared = c(dt_r2_train, dt_r2_test, rf_r2_train, rf_r2_test, sgb_r2_train, sgb_r2_test, xgb_r2_train, xgb_r2_test)
)

# Render the performance comparison table with blue and green colors
performance_comparison %>%
  kable("html", caption = "Performance Comparison of Different Models") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = F) %>%
  column_spec(1, bold = TRUE, background = "#CCE5FF") %>%  # Light blue for Model column
  column_spec(2, bold = TRUE, background = "#D5F5E3") %>%  # Light green for Dataset column
  column_spec(3, background = "#D5E4EB") %>%  # Light blue for RMSE column
  column_spec(4, background = "#D5E4EB") %>%  # Light blue for MAE column
  column_spec(5, background = "#D5E4EB") %>%  # Light blue for R_squared column
  row_spec(0, bold = TRUE, color = "white", background = "darkblue")


```


##### The best model to choose is the XGboost for its superior performance on the test set because the ability of easily addressing overfitting. This model will likely perform best on unseen data.



```{r}
models <- list(
  Model = c("Decision Tree","Random Forest", "Stochastic Gradient Boosting", "XGBoost"),
  `Training Performance` = c("Poor","Good", "Moderate", "Very Good"),
  `Test Performance` = c("Very poor","Good (Similar to Training)", "Moderate (Slight Drop)", "Excellent (Large Drop)"),
  `Overfitting Indication` = c("Low","Low", "Potential", "High")
)

model_data <- data.frame(models)

cat("Model", " ", "Training Performance", " ", "Test Performance", " ", "Overfitting Indication", sep = "\t", fill = TRUE)
cat("\n")
print(model_data, rownames = FALSE)

```



